Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
Another rewrite of the abstract

\newif\ifdraft\drafttrue  % set true to show comments

@@

{\em Lenses}~\cite{Lenses} are programs that specify bidirectional
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.
%
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses
where the two data formats are considered modulo some equivalence relation.
The use of quotients allows the programmer to ignore inessential differences
in the data, such as the amount of whitespace or the ordering of fields.

Quotient lenses can be difficult to program, requiring the programmer to
give both a specification for the data on each side of the lens and a
separate specification for the equivalence relations on the data.  The
programmer must also manually implement a lens program that maps between the
formats while respecting the equivalence relations---this is itself a
challenging task thanks to the restrictions on how lenses can be combined
while still satisfying the lens laws.  \bcp{Don't really understand that
  paragraph!}

This paper addresses both of these challenges\bcp{what are the two
  challenges?}. First, we propose the language of {\em Quotient Regular
  Expressions}, or QREs, which define a broad class of equivalence relations
directly on regular languages using a compact and convenient notation. This
eliminates the need to specify the data and the equivalence relation on the
data separately. We then define {\em QRE lenses}, a class of quotient lenses
that are typed using QREs, and show how to {\em synthesize} QRE lenses from
their QRE types and a set of input-output examples. Finally, we demonstrate
the applicability of this approach by synthesizing QRE lenses between
several real-world data formats from the {\tt data.gov} database.
\end{abstract}

\section{Introduction}
\bcp{Not sure there is time to fix this, but the intro seems very long and
  technical---it takes a long time to get to saying what our actual
  contributions are!  Wonder if we should move most of it to a second
  ``Example'' section.} 
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

>>>>

\newif\ifdraft\drafttrue  % set true to show comments
\newif\ifplentyoftime\plentyoftimefalse  % :-)

@@

{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.  
%
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses
where each of the data formats is quotiented by some equivalence relation
and the lens laws are taken modulo equivalence.  This allows lenses to be
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

\smallskip

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer must
separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these
three share a great deal of structure.  Moreover, writing the lens between
the equivalence classes can be challenging due to the rather fiddly, mainly
due to ``non-ambiguity'' conditions in the lens well-formedness rules.

\smallskip

In this paper, we address both of these issues.  First, we propose a novel
language of {\em Quotient Regular Expressions}, or QREs, which allows
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular
language and its equivalence classes are captured by a single expression.
We then define {\em QRE lenses}, a class of bijective quotient lenses that
are typed using QREs, and show how to {\em synthesize} QRE lenses from their
QRE types and a set of input-output examples.  We demonstrate the
practicality of this approach by synthesizing QRE lenses between several
real-world data formats from the {\tt data.gov} database.
\end{abstract}

\section{Introduction}
\ifplentyoftime
\bcp{Not sure there is time to fix this, but the intro seems very long and
  technical---it takes a long time to get to saying what our actual
  contributions are!  Wonder if we should move most of it to a second
  ``Example'' section.} 
\fi
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}\bcp{there are many others; we
  should mention all of them} enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}
